Perihelion advance of Mercury

### 2019-06-17 perihelion_advance_of_mercury.Rmd
typos and clarification

The orbit of Mercury has long been known to advance at a rate of about
575'' [seconds of arc] per century.  Various classical phenomena, such
as the effects of Venus, and the oblateness of the sun, account for
all but 43'' and this remaining amount puzzled astronomers for many
years.  According to Einstien and others, the general theory of
relativity predicts an additional precession of almost exactly 43
seconds per century, thus closing the gap between expectation and
observation.  In this document I will calculate the precessional rate
using numerical methods.

@@

```{r}
jj <- trajectory(30, 1/2, 3) # m=1/2 -> Schwarzschild radius = 1
plot(jj$x,jj$y, asp=1, type='l', axes=FALSE, xlab='', ylab='',main='Geodesics in the Schwarzschild geometry')
polargrid(r=10*(1:3))
event_horizon()
```

In the above, we see a typical orbit in Schwarzschild space and a precession of around 40 degrees per orbit.  As a consistency check we can perform the same calculation but with Newtonian approximations:

@@

and above we see that the orbit is accurately closed, as expected.

@@

First we specify some orbital parameters for the orbit of Mercury, mostly taken from wikipedia:
```{r}
G <-  6.674e-11     # big G in SI
M  <- 1.9885e30     # mass of Sun (kg)
M_merc <- 3.3011e23 # mass of Mercury (kg)
sol <- 299792458    # speed of light (m/s)
merc_perihelion <- 46001200e3 
merc_aphelion   <- 69816900e3
mu <- G*(M + M_merc)
```

@@

And some futher features using classical geometry:
```{r}
a <- (merc_perihelion + merc_aphelion)/2
cee <- a-merc_perihelion
b <- sqrt(a^2-cee^2)
ecc <- sqrt(1-(b/a)^2)  # should match ecc_wiki above
ell <- (a^2-cee^2)/a  # semi-latus rectum
```

@@

With all these we can simulate the orbit:
```{r}
 out <- trajectory(merc_perihelion, m=d, h=h, tau=seq(from=0,to=35e15,len=5e5))
 plot(out$x,out$y,type='l',asp=1,axes=FALSE,xlab='',ylab='')
 event_horizon()
 max(out$phi/(2*pi))
```

In the above diagram we see over 15.4 orbits of Mercury, exhibiting no
visible deviation from an exactly closed orbital path.  To calculate
the precession we need to simulate the motion for a longer time:

@@

plot(tee,phi-pi)
```

There is a lot of noise on this caused by the discrete nature of the simulation results provided by ```ode()```.  To estimate the precession we can use linear regression:

@@

To convert this figure which is given in radians per orbit) to
arcseconds per century, we need to multiply by a sequence of factors:

>>>>

The orbit of Mercury has long been known to advance at a rate of about
575'' [seconds of arc] per century.  Various classical phenomena, such
as the effects of Venus, and the oblateness of the sun, account for
all but 43'' and this remaining amount puzzled astronomers for many
years.  According to Einstein and others, the general theory of
relativity predicts an additional precession of almost exactly 43
seconds per century, thus closing the gap between expectation and
observation.  In this document I will calculate the precessional rate
using numerical methods.

@@

```{r}
# Simulate a trajectory starting at r=30, mass of sun=1/2 so Schwarzschild radius=1:
jj <- trajectory(r_start=30, m=1/2, h=3) # h=3 gives angular momentum
plot(jj$x,jj$y, asp=1, type='l', axes=FALSE, xlab='', ylab='',main='Geodesics in the Schwarzschild geometry')
polargrid(r=10*(1:3))
event_horizon()  # show the event horizon, to scale
```

(NB ```trajectory()``` takes geometrized units).  In the above, we see
a typical trajectory in Schwarzschild space with a precession of
around 40 degrees per orbit.  As a consistency check we can perform
the same calculation but with Newtonian approximations:

@@

and above we see that the orbit is accurately closed, as expected
(three orbits are simulated, the lines overlap).

@@

First we specify some orbital parameters for the orbit of Mercury, mostly taken from Wikipedia:
```{r}
G <-  6.674e-11     # big G in SI
M  <- 1.9885e30     # mass of Sun (kg)
M_merc <- 3.3011e23 # mass of Mercury (kg)
sol <- 299792458    # speed of light (m/s)
merc_perihelion <- 46001200e3 
merc_aphelion   <- 69816900e3
mu <- G*(M + M_merc)
```

@@

And some further features using classical geometry:
```{r}
a <- (merc_perihelion + merc_aphelion)/2
cee <- a-merc_perihelion
b <- sqrt(a^2-cee^2)
ecc <- sqrt(1-(b/a)^2)  # should match eccentricity given by Wikipedia
ell <- (a^2-cee^2)/a  # semi-latus rectum
```

@@

With all these we can simulate the orbit:
```{r}
 out <- trajectory(merc_perihelion, m=d, h=h, tau=seq(from=0,to=35e15,len=5e5))
 plot(out$x,out$y,type='l',asp=1,axes=FALSE,xlab='',ylab='',main='Orbit of Mercury')
 event_horizon()
 max(out$phi/(2*pi))
```

In the above diagram we see over 15.4 orbits of Mercury, exhibiting no
visible deviation from an exactly closed orbital path.  Nevertheless,
the orbit is not closed as Mercury has a tiny amount of orbital
precession.  To calculate this precession we need to simulate the
motion for a longer time:

@@

plot(tee,phi-pi,main='orbital angle at aphelion',xlab='time (seconds)',ylab='angle of aphelion')
```

The above diagram shows a clear trend, somewhat obscured by a vertical
spread due to the discrete nature of the simulation.  To estimate the
underlying trend, which is the general relativistic precession of
Mercury, we can use linear regression:

@@

To convert this figure (which is given in radians per orbit) to the
more usual units of arcseconds per century, we need to multiply by a
sequence of factors:
